Dynamic Stochastic Approximation Monte Carlo

### 2018-05-21 papers/histogram/sad.tex
added some text on Tmin and fixed some SAD formulae.

@@

\section{SAD Algorithm}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. While for
SAMC the update factor is defined in the original
implementation~\cite{liang2007stochastic} in terms of an arbitrary
tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{t_0}{\max(t_0,t)}
\end{align}
The update factor $\gamma_{t}^{\text{SA}}$ can be thought of as
$\nicefrac{\text{d}S}{\text{d}t}$. This tells us that $t_0$ should have
dimensions of entropy.  We argue that we can write this entropy `term'
geometrically as an area bounded by the $\ln{D(E)}$ i.e. the entropy.  This
gives the following
\begin{align}
\gamma_{t}^{\text{SA}} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas bounded
by the $\ln{D(E)}$ and the minimum temperature line.
The update factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course
of the simulation, the update factor experiences a jump. This behavior allows
SAD to dynamically prevent the update factor from decreasing too rapidly.

Fig.~\ref{fig:gamma-vs-t}. shows the monotonic decrease of the update
factor for SAMC.  The update factor for WL decreases like a step
function while the update factor for SAD dynamically fluctuates around
a value less than 1 for early MC moves (while finding new energy
states).  Ultimately once all energy states have been found, the update
factor for SAD decreases in the same monotonic fashion as SAMC.

\subsubsection{High and Low energy racheting}
Since SAD explores all energy states it is imperative that the method have
a reasonable idea about where the minimum important energy and maximum
entropy state are located in energy space. Since these are not user input
parameters like for methods suchas WL-TMMC, the simulation is responsible
for determining and updating these energy locations. In order to prevent,
the method from experiencing errors near the energy extrema, we employ a 
high energy and a low energy to define the range over which the energy 
histogram is made flat. The high energy is the highest ever value that the
maximum entropy state has taken while the low energy is the lowest ever value
that the minimum important energy has taken. Together these two energies serve
as a kind of ratcheting effect to determine the energy landscape boundaries.

>>>>

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/N500-lndos-comparison}
  \caption{Put this in the right place}
\end{figure}

@@

\section{SAD Algorithm}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. Each
Broad-Histogram method has unique advantages and disadvantages.
Wang-Landau requires an energy range for initialization.  SAMC removes
this energy range requirement but requires simulating every possible
energy. Our proposed method SAD requires the user to input
$T_\text{min}$ which is an immediate disadvantage of the method;
however, this allows the simulating to define an energy simulation
range between $T_\text{min}$ and $\infty$. Setting a physical parameter
such as a minimum temperature $T_\text{min}$ is intuitive vs. a user
needing to know a prior either an energy range or some unphysical
parameter $t_0$.

While for SAMC the update factor is defined in the original
implementation, for SAD the update factor $\gamma_{t}^{\text{SA}}$ can
be thought of as $\nicefrac{\text{d}S}{\text{d}t}$. This tells us that
$t_0$ should have dimensions of entropy.  We argue that we can write
this entropy `term' geometrically as an area bounded by the $\ln{D(E)}$
i.e. the entropy. This gives the following
\begin{align}
\gamma_{t}^{\text{SA}} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state.  The factor of $3$ comes from the ratio of areas
bounded by the $\ln{D(E)}$ and the minimum temperature line. The update
factor has the same $1/t$ behavior as the original SAMC algorithm;
however now every time a new energy is discovered during the course of
the simulation, the update factor experiences a jump. This behavior
allows SAD to dynamically prevent the update factor from decreasing too
rapidly. In contrast to SAD, Fig.~\ref{fig:gamma-vs-t}. shows the
monotonic decrease of the update factor for SAMC.  The update factor
for WL decreases like a step function while the update factor for SAD
dynamically fluctuates around a value less than 1 for early MC moves
(while finding new energy states).  Ultimately once all energy states
have been found, the update factor for SAD decreases in the same
monotonic fashion as SAMC.

\subsubsection{High and Low energy racheting}
Since SAD explores all energy states it is imperative that the method
have a reasonable idea about where the minimum important energy and
maximum entropy state are located in energy space. Since these are not
user input parameters such as they are for methods like WL-TMMC, the
simulation is responsible for determining and updating these energy
locations. In order to prevent, the method from experiencing errors
near the energy extrema, we employ a high energy $\epsilon_H$ and a low
energy $\epsilon_L$ to define the range over which the energy histogram
is made flat. $\epsilon_H$ is the highest ever value that the maximum
entropy state has taken while $\epsilon_L$ is the lowest ever value
that the minimum important energy has taken. SAD uses these two
energies to control the weights $w(E)$ above $\epsilon_H$ and
$\epsilon_L$. If the energy $E$ is greater than $\epsilon_H$ then the
weights can be calculated as ~\jpsays{equation formatting problems!}
\begin{equation}
\begin{aligned}
\ln w(E) = \max(\ln w(E) + \gamma, \\
    \ln w(\epsilon_H) + \log(e^{w(E)-\ln w(\epsilon_H)}))
\end{aligned}
\end{equation}
If the energy $E$ is less than $\epsilon_L$ then the weights can be
calculated as
\begin{equation}
\begin{aligned}
\ln w(E) = \max(\ln w(E) + \gamma, \ln w(\epsilon_L) \\
    + \frac{\epsilon_L - E}{T_\text{min}}
    + \log(\gamma + e^{w(E) -\ln w(\epsilon_L)
    + \frac{\epsilon_L -E}{T_\text{min}}}))
\end{aligned}
\end{equation}
Sad sets the maximum entropy state equal to the energy if either of the
following is true.
\begin{align}
\ln w(E) > \ln w(S_\text{max}) \; \text{or} \; H(S_\text{max}) = 0
\end{align}
Thus if the method is beyond the maximum entropy state or has collected no
statistics at the maximum entropy state then it is set equal to the energy.
Similarly, Sad sets the minimum important energy equal to the energy if
either of the following is true.
~\jpsays{equation formatting problems!}
\begin{align}
\ln w(E) > \ln w(E_\text{min}) - \frac{E_\text{min}-E}{T_\text{min}}
\; \text{or} \; H(E_\text{min}) = 0
\end{align}
Thus if the method is beyond the minimum important energy or has collected no
statistics at the minimum important energy then it is set equal to the energy.
Together these two energies serve as a kind of ratcheting effect to
determine the energy landscape boundaries. Outside the range set by the
weights $w(E)$, the maximum entropy state is set to the energy $E$. The
minimum important energy is also set to the energy when outside the
calculated range.
